\chapter{Sample annotations}
\label{app:sample-annotations}

This appendix contains small sample annotations from the English and Nuuchahnulth corpora. Each token that was included in this study is given an annotation indicating its discourse function—\textsc{ref}, \textsc{pred}, or \textsc{mod} for reference, predication, or modification, respectively. Lexical items that are excluded from this study for one of the reasons discussed in \secref*{sec:3.3} are not given an annotation. Recall from \chref{ch:methods} that for English, predicate adjectives and non-finite verbs are \emph{not} included. For English, the annotation is given as a subscript after the word. For Nuuchahnulth, the discourse function is indicated as a \nth{5} line in the interlinear gloss.

\section{English}
\label{app:sample-annotations-English}

\onehalfspacing

Well life\func{ref} there in the country\func{ref} is nice and tranquil. I lived\func{pred} working\func{mod} all of my life\func{ref} with livestock\func{ref}. I always had to get\func{pred} up early milk\func{pred} the cows\func{ref} and uh run\func{pred} run\func{pred} them as we say\func{pred} because it's a— to the pastures\func{ref} until times\func{ref} got\func{pred} pretty bad and one\func{mod} day\func{ref} I sent\func{pred} my daughter\func{ref} to to the pasture\func{ref} to bring in the cows\func{ref}. We brought\func{pred} them back in the afternoon\func{ref} when I saw\func{pred} that behind her there came\func{pred} a big\func{mod} group\func{ref} of they looked\func{pred} like soldiers\func{ref} but in street\func{mod} clothes\func{ref}. Then she came\func{pred} my daughter\func{ref} came\func{pred} almost green pale and she said\func{pred} to me "Mama" she said\func{pred} to me "Those are guerillas\func{ref}!" That was the first\func{mod} time\func{ref} I saw\func{pred} them the gue— the guerillas\func{ref}. \parencite[ArguetaBertila-ENG]{OANC}

\section{Nuuchahnulth}
\label{app:sample-annotations-Nuuchahnulth}

\singlespacing
\renewcommand{\eachwordfive}{\rule[-10pt]{0pt}{0pt}\rmfamily}

\glllll qiiʔaƛ                   qiit̓an̓aƛ                          qʷiyuckʷiʔitq        qʷis        ʔaḥ\\
        qiː‑ʼaƛ                  qiː‑t̓an̓a‑ʼaƛ                      qʷiyu‑ckʷi·‑ʔi·tq    qʷis        ʔaḥ\\
        for.a.long.time‑\gl{fin} for.a.long.time‑slightly‑\gl{fin} time‑done‑\gl{rel.3} happen.thus this\\
        happened.long.ago        quite.a.while.ago                 when.it.occurred     happen.thus this\\
        \gl{pred}                \gl{pred}                         \gl{ref}             \gl{pred}   \gl{ref}\\
        \vfix
        \tln{This happened a long time ago.}

\glllll siikc̓inƛ           siikaa            hitac̓inƛ                  maaqtusiis\\
        siːk‑c̓inƛ          siːk‑(y)a·        hita‑c̓inƛ                 maːqtusiːs\\
        sailing‑into.a.bay sailing‑\gl{cont} there.\gl{mom}‑into.a.bay \gl{name}\\
        sail.into.a.bay    sailing           entered.into.a.bay        \gl{name}\\
        \gl{pred}          \gl{pred}         \gl{pred}                 \gl{ref}\\
        \vfix
        \tln{They sailed into the bay of Maaqtusiis.}

\glllll yuupickʷimatak              yuupi    yuksaaʔa\\
        yuːpi‑ckʷi·‑matak           yuːpi    yu‑ksa·ʔa\\
        breeze‑done‑probably        breeze   blowing‑come.to.land\\
        probably.there.was.a.breeze breeze   breeze.along.the.shoreline\\
        \gl{pred}                   \gl{ref} \gl{pred}\\
        \vfix
        \tln{There probably was a little wind, blowing towards the land.}

\glllll qʷiyimtii                   n̓aas     hitac̓inƛ\\
        qʷiyu‑imt‑(y)iː             n̓aːs     hita‑c̓inƛ\\
        when‑\gl{past}‑\gl{indef.3} day      there.\gl{mom}‑in.a.bay\\
        whenever.it.was             day      entered.a.bay\\
        \gl{pred}                   \gl{ref} \gl{pred}\\
        \vfix
        \tln{They came into the bay one day.}

\glllll hitac̓u      ʔukɬaakʔakna                          y̓uuqʷaa\\
        hitac̓u      ʔu‑kɬa·‑ak‑ʔa·k‑na·                   y̓uːqʷaː\\
        \gl{name}   it‑called‑\gl{dur}‑\gl{poss}‑\gl{1pl} also\\
        \gl{name}   we.also.call.it                       also\\
        { }         \gl{pred}                             \gl{pred}\\
        \vfix
        \tln{We also call it (the bay) ``\txn{hitac̓u}".}

\glllll waɬyuu       maaqtusiis  wiiḥaaqsusiis\\
        waɬ‑yu·      maːqtusiːs  wiːḥaːqsusiːs\\
        go.home‑done \gl{name}   \gl{name}\\
        gone.home    \gl{name}   \gl{name}\\
        \gl{pred}    \gl{ref}    { }\\
        \vfix
        \tln{They went to Maaqtusiis — [to be exact,] \txn{Wiiḥaaqsusiis}.}

\glllll ʔuʔiiyačištckʷi          wiiḥaaqsusiis t̓ayuukʷiƛ         kuunaa\\
        ʔu‑ʔi·ya‑ačišt‑ckʷi·     wiːḥaːqsusiːs t̓ayuː‑kʷi(ƛ)      kuːnaː\\
        it‑reach‑on.the.sea‑done \gl{name}     anchored‑\gl{mom} schooner\\
        reached                  \gl{name}     anchored          schooner\\
        \gl{pred}                \gl{ref}      \gl{pred}         \gl{ref}\\
        \vfix
        \tln{The schooner reached \txn{Wiiḥaaqsusiis} and dropped anchor.}

\glllll wik       ʔiiḥ      wikckʷii  ʔiiḥ\\
        wik       ʔiːḥ      wik‑ckʷi· ʔiːḥ\\
        not       large     not‑done  large\\
        not       large     was.not   large\\
        \gl{pred} \gl{pred} \gl{pred} \gl{pred}\\
        \vfix
        \tln{It (the schooner) was not so big.}

\glllll ʔaƛa      ʔaƛista             qacc̓istamitquu\\
        ʔaƛa      ʔaƛa‑ista           qacc̓a‑ista‑mit‑quː\\
        two       two‑people.on.board three‑people.on.board‑\gl{past}‑\gl{cond.3}\\
        two       two.people.on.board there.could.have.been.three.people.on.board\\
        \gl{pred} \gl{ref}            \gl{pred}\\
        \vfix
        \tln{There were two crewmen, or there could have been three, on the ship.}

\glllll hinaačiƛ̓aɬ                            yaqitii\\
        hin‑a·či(ƛ)‑ʔa·ɬ                      yaq‑it‑(y)iː\\
        there.\gl{mom}‑go.out.to.meet‑\gl{pl} who‑\gl{past}‑\gl{indef.3}\\
        they.go.out.to.meet                   whoever.it.was\\
        \gl{pred}                             \gl{ref}\\
        \vfix
        \tln{Some people went out to meet them (the people on the schooner).}

\glllll ʔin   ʔutwiickʷiʔaaɬ          hinaačiƛ                      wiʔakʔi        wiiʔaksaʔi           ḥaaʔak̓atʔi                ɬim̓aqsti\\
        ʔin   ʔutwiː‑ckʷi·‑ʔaːɬ       hin‑a·či(ƛ)                   wiʔak‑ʔi·      wiʔak‑sa‑ʔi·         ḥaːʔak‑ʼat‑ʔi·            ɬim̓aqsti\\
        since first‑done‑always       there.\gl{mom}‑go.out.to.meet brave‑\gl{def} brave‑real‑\gl{def}  strong‑\gl{poss}‑\gl{def} mind\\
        since they.were.the.first.one go.out.to.meet                the.brave.one  the.bravest.one      the.one.with.strong.one   mind\\
        { }   \gl{pred}               \gl{pred}                     \gl{ref}       \gl{ref}             \gl{ref}                  \gl{ref}\\
        \vfix
        \tln{The first ones to go out were the bravest ones, the ones with strong minds.}

\glllll ʔin naʔaackʷaƛ         ʔaya     mamaɬn̓i   hisiickʷiʔitqʔaɬ           hiistiƛ   ciqy̓ak           čiinuukʔatḥ\\
        ʔin naʔaː‑ckʷi·‑ʼaƛ    ʔaya     mamaɬn̓i   hisiː‑ckʷi·‑ʔi·tq‑ʔa·ɬ     hiːstiƛ   ciq‑y̓akʷ         č̓iːnuːk‑ʼatḥ\\
        ??  hear‑done‑\gl{fin} many     white.man ??‑done‑\gl{rel.3}‑\gl{pl} from      speak‑instrument Chinook‑belonging.to\\
        ??  understood         many     white.man the.way.they.spoke         from      language         Chinook\\
        { } \gl{pred}          \gl{mod} \gl{ref}  \gl{ref}                   \gl{pred} \gl{ref}         \gl{pred}\\
        \vfix
        \tln{Many white men could understand Chinook Jargon.}

\glllll čiičiinukʷackʷaƛ                                ʔuuš\\
        \gl{dup}‑čiːnu·k‑(y)a‑ckʷi·‑ʼaƛ                 ʔuːš\\
        \gl{distr}‑speak.Chinook‑\gl{rep}‑done‑\gl{fin} some\\
        spoke.Chinook.Jargon                            some\\
        \gl{pred}                                       \gl{ref}\\
        \vfix
        \tln{Some of them spoke Chinook Jargon.}

\glllll hist̓atḥckʷaƛukʔaɬ                                  ʔaḥ      [Hudson Bay]\\
        hist‑ʼatḥ‑ckʷi·‑ʼaƛ‑uk‑ʔa·ɬ                        ʔaḥ      ~\\
        there‑belonging.to‑done‑\gl{fin}‑\gl{poss}‑\gl{pl} this     ~\\
        they.got.theirs.from.there                         they     ~\\
        \gl{pred}                                          \gl{ref} { }\\
        \vfix
        \tln{They got theirs (= their knowledge of Chinook Jargon) from Hudson Bay Company.}

\glllll yaqʷiiyii               naʔaaʔaƛ      Captainmitquu                 yaqʷacʔitq                  šipʔii\\
        yaqʷ‑wi·‑(y)iː          naʔaː‑ʼaƛ     Captain‑mit‑quː               yaqʷ‑ac‑ʔi·tq               šip‑ʔi·\\
        who‑first‑\gl{indef.3}  hear‑\gl{fin} captain‑\gl{past}‑\gl{cond.3} who‑belonging.to‑\gl{rel.3} ship‑\gl{def}\\
        the.ones.who.were.first understood    one.who.was.Captain           owner.of                    the.ship\\
        \gl{ref}                \gl{pred}     \gl{pred}                     \gl{ref}                    \gl{ref}\\
        \vfix
        \tln{Among the first ones that [learned to] understand the language might have been the Captain who was taking command of the ship.}

\renewcommand{\eachwordfour}{\rmfamily}

\parencite[Kingfisher]{Louie2003}
