\clearpage
\phantomsection
\section*{Acknowledgments}
\label{sec:acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}

\singlespacing

So many people have aided me in completing this dissertation—either directly or indirectly, in ways big and small—that I cannot possibly thank them all here. To those I have omitted: Thank you. You are appreciated.

For sparking my original interest in lexical flexibility/polyfunctionality, I thank Drs. Elaine Francis (Purdue University) and Greville G. Corbett (University of Surrey), both of whom I had the privilege of taking courses from at the 2011 Linguistic Institute at the University of Colorado, Boulder.

For financial support at different stages during this project, I would like to thank the National Science Foundation's Graduate Research Fellowship Program, the University of California, Santa Barbara, and the Department of Linguistics there. This project would not have been possible without their support.

For kindly providing me with the corpus of Nuuchahnulth, my sincere thanks go to Dr. Toshihide Nakayama (Tokyo University of Foreign Studies). The task of preparing a documentary corpus often goes underappreciated, but is in fact one of the most theoretically and analytically taxing tasks a linguist can undertake, requiring a broad understanding of many different subfields of linguistics. Thank you for your many years of meticulous work that made this project possible.

For providing me the opportunity to present an early version of this work, I would like to thank Dr. Anya Lunden, Dr. Jack B. Martin, and the linguistics department at the College of William \& Mary.

I have benefitted from discussions on lexical polyfunctionality, and gotten feedback on this project, from many people. Thank you especially to Dr. Eric R. Hansen (Loyola University Chicago), Dr. Hunter T. Lockwood (Myaamia Center), Dr. Monica Macaulay (University of Wisconsin-Madison), Dr. Jack B. Martin (College of William \& Mary), Phillip G. Rogers (University of California, Santa Barbara), and Dr. Sandra A. Thompson (University of California, Santa Barbara) for these discussions.

A special thanks is also due to my dissertation committee: Drs. Bernard Comrie, Stefan Th. Gries, and Marianne Mithun at the University of California, Santa Barbara, and Dr. William Croft at the University of New Mexico. Your patience with my much-delayed research process, the time you've dedicated to advising me and providing feedback, and your insightful comments throughout this project are all greatly appreciated.

To all the faculty at the University of California, Santa Barbara linguistics department: Thank you. I could not have asked for a better intellectual home and community of scholars.

Very special thank yous are also due to some of my mentors. Marion Bittinger (Rosetta Stone) patiently helped teach me to become a professional, and provided me the opportunity to work with Native communities on language revitalization for the first time. Jack Martin (College of William \& Mary) offered me invaluable professional advice throughout my career, both in industry and academia, and always took an interest in whatever I was working on. Carol Genetti (New York University Abu Dhabi) took a holistic approach to advising and taught me the importance of self-care, and supported my interest in East African languages in numerous ways. Without each of your friendships and guidance I would not be where I am today.

A number of people have shown incredible patience with me while I worked to complete this project, putting aside some of their own goals and projects temporarily so that I could complete mine. To Antti Arppe, Erin Daniels, Jack DuBois, Monica Macaulay, Marianne Mithun, and Kimberly Walden, you have my deepest gratitude.

The degree of personal support I have received during my Ph.D. and this dissertation project is overwhelming and humbling. To Andrew Grubb†, Bobby López, Brendon Yoder, Eric Hansen, Erin Daniels, Jack Martin, James Oren, Jessi Love-Nichols, Kim Walden, Maggie Grubb-Hassett, Morgan Sleeper, and Stephen Boatwright, I cannot thank you enough. You have each made me better in your own way.

To Patrick and Phill: I am so glad we got to take our Ph.D. journeys together. Thank you for being there every step of the way.

To Marianne: It has been one of the greatest privileges of my life to study with you. Every conversation with you reminds me of the wonder in what we get to research and discover. You continued to believe in me and support me when things got tough both academically and personally. Thank you for helping keep my spark for language alive.

To my family: You've sacrificed so much for me so that this could be possible. You were there for me even in times when I didn't know how to be there in return. I am grateful for each of you every single day. I love you.

\hfill\textasciitilde{ }Danny

\doublespacing
