\newlength{\datewidth}
\setlength{\datewidth}{6em}

\renewcommand{\arraystretch}{1.5}

\newenvironment{cvsection}[1]
{
  \subsection*{#1}
  \begin{longtabu} to \textwidth { p{\datewidth} >{\hangindent=\defaulthang}X[l] }
}
{\end{longtabu}}

\singlespacing

\section*{Curriculum Vitae}
\label{sec:cv}
\addcontentsline{toc}{section}{Curriculum Vitae}

\begin{center}
  \theauthor
\end{center}

\begin{cvsection}{Education}
  December 2020 & Ph.D. in Linguistics, University of California, Santa Barbara\\
  March 2016    & M.A. in Linguistics, University of California, Santa Barbara\\
  June 2008     & B.A. in Linguistics \& Philosophy, The College of William \& Mary (magna cum laude)
\end{cvsection}

\begin{cvsection}{Professional Experience}
  2018--2019 & Editor, Custom Language Products, Rosetta Stone\\
  2015--2017 & Teaching Assistant, Department of Linguistics, University of California, Santa Barbara\\
  2014--2015 & Research Assistant (under Prof. Carol Genetti), Department of Linguistics, University of California, Santa Barbara\\
  2011--2013 & Associate Researcher, Research Labs, Rosetta Stone\\
  2008--2011 & Editor, Endangered Languages Program, Rosetta Stone\\
  2007--2008 & Intern, Endangered Languages Program, Rosetta Stone\\
  2006       & Spanish Instructor, Nielsen Builders\\
  2004--2006 & Lab Assistant, Language Lab, The College of William \& Mary\\
  2003--2004 & Latin Instructor, Bridgewater Home School Unit
\end{cvsection}

\begin{cvsection}{Publications}

  2019 & The Chitimacha language: A history. In Nathalie Dajko \& Shana Walton (eds.), \pubtitle{Languages in Louisiana: Community \& culture} (America's Third Coast Series). University Press of Mississippi.\\

  2019 & Semantic alignment in Chitimacha. \pubtitle{International Journal of American Linguistics} 85(3): 313--363. doi:\href{https://doi.org//10.1086/703239}{10.1086/703239}.\\

  2018 & Category genesis in Chitimacha: A constructional approach. In Kristel Van Goethem, Muriel Norde, Evie Coussé, \& Gudrun Vanderbauwhede (eds.), \pubtitle{Category change from a constructional perspective} (Constructional Approaches to Language 20), 15--46. John Benjamins. doi:\href{https://doi.org/10.1075/cal.20.02hie}{10.1075/cal.20.02hie}.\\

  2016 & \pubtitle{The cohesive function of prosody in Ékegusií (Kisii) narratives: A functional-typological approach}. M.A. thesis, University of California, Santa Barbara.\\

  2013 & On linguistics, language, and our times: A linguist's narrative reviewed. \pubtitle{Linguistic Typology} 17(2): 291--321. Review article of \pubtitle{I am a linguist} by R. M. W. Dixon (Brill, 2010). doi:\href{https://doi.org/10.13140/RG.2.2.13238.96329}{10.13140/RG.2.2.13238.96329}.\\

  2013 & (with Sharon Hargus \& Edward Vajda, eds.) \pubtitle{Working papers in Athabaskan (Dene) languages 2012}. Alaska Native Language Center Working Papers 11. ANLC.

\end{cvsection}

\begin{cvsection}{Awards}
  2019 & SSILA Best Student Presentation Award\\
  2015 & National Science Foundation (NSF) Graduate Student Research Fellowship (GRFP)\\
  2015 & \nth{2} place, University of California Grad Slam\\
  2015 & Winner, University of California, Santa Barbara Grad Slam\\
  2013 & Chancellor's Fellowship, University of California, Santa Barbara\\
  2006 & Boren Scholarship, National Security Education Program (NSEP)
\end{cvsection}

\subsection*{Fields of Study}

\begin{longtabu} to \textwidth {>{\hangindent=\defaulthang}X[l]}

  Major Fields: Linguistic Typology, Language Documentation \& Description, Language Revitalization, Prosody, Discourse, Language Change, Language Contact, Digital Linguistics, Corpus Linguistics\\

  Linguistic Typology with Professor Bernard Comrie \& Professor Marianne Mithun\\

  Language Documentation \& Description with Professor Eric Campbell, Professor Carol Genetti, \& Professor Marianne Mithun\\

  Language Revitalization with Professor Carol Genetti\\

  Prosody with Professor Carol Genetti, Professor Matthew Gordon, \& Professor Marianne Mithun\\

  Discourse with Professor Patricia Clancy, Professor John W. DuBois, Professor Carol Genetti, \& Professor Marianne Mithun\\

  Language Change with Professor Marianne Mithun\\

  Language Contact with Professor Marianne Mithun\\

  Digital Linguistics with Professor Eric Campbell \& Professor Stefan Th. Gries\\

  Corpus Linguistics with Professor Eric Campbell, Professor John W. DuBois, \& Professor Stefan Th. Gries

\end{longtabu}

\doublespacing
